NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2018-06-19 Manuscript/manuscriptPS.tex
Update text

\subsection{Counterion binding to lipid bilayers containing PS lipids}
Since systems with negatively charged PS lipids always contain counterions
and the ion binding presumably affects the PS headgroup
order parameters, the discussion of ion binding affinity and
headgroup structure cannot be separated as done previously
for PC lipids \cite{botan15,catte16}. The density profiles
of counterions along membrane normal from different simulations
show significant binding of counterions in bilayers in all
models in Fig. \ref{NAdensPOPS}.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/NAdensPOPS.eps}
  \caption{\label{NAdensPOPS}
    Counterion densities of POPS lipid bilayer along the membrane normal from
    simulations with different force fields.
  }
\end{figure}
Differences are, however,
observed in the total binding affinity and in the shape of density profiles.
Comparison of bulk concentrations suggests that strongest binding
is observed in MacRog, Berger and Lipid17 models, respectively.
Other models gives roughly similar total binding affinity
based on rough comparison of bulk concentrations.
In CHARMM36, CHARMM36ua and Gromos-CKP models two maxima are observed
in sodium density plots, possible corresponding binding to phoshate
and carboxylic grops in the PS headgroup. In other models only
a single maxima is observed. \todo{More quantitative analysis of
binding positions and affinity may be useful.}

Based on previous simulation results it has been suggested that
the counterion binding and concomitant screening of headgroup repulsion
leads to smaller area per lipid in PS lipid bilayers than in PC
bilayers \cite{pandit02,mukhopadhyay04,pedersen06}. Therefore we also show the area per lipid from different models
in Fig. \ref{NAdensPOPS}. The proposed trend is observed for Lipid17 model,
for which the Joung-Cheatham ions give higher affinity and smaller area
per molecule. However, the trend is not observed when comparing accross different
force fields. For example, CHARMM36ua and MacRog give similar area per molecule
but binding affinity is significantly higher in MacRog. On the other hand,
Gromos-CKP models gives a larger area per lipid than other models, being only
in agreement with experiments \cite{pan14}, but the binding affinity is comparable
to the CHARMM36, CHARMM36us and Slipids, which give smaller are per molecule.

In order to compare the sodium interaction with PS containing membranes between
experiments and simulations, Fig. \ref{PSresponseTONaCl} shows the headgroup
order parameters of POPC and POPS measured from 5:1 mixture as a
function of different monovalent ions added to the buffer.
The experimental experimental results \cite{roux86} are compared with different simulation
models.
\begin{figure*}[]
  \centering
  \includegraphics[width=18.0cm]{../Figs/CHANGESwithMONVALENTwithPS.eps}
  \caption{\label{PSresponseTONaCl}
    Changes of the PC (left) and PS (right) headgroup order parameters as a function of
    added NaCl, KCl and LiCl from POPC:POPS (5:1) mixture. The experimental data is from Ref. \citenum{roux90}.
    The values from counterion-only systems are set as a zero point of y-axis.
    To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
    in PS headgroup (bottom, right), the y-axis is tranferred with the same value for both order parameters such that the lower order
    parameter value is at zero.
  }
  \todo{The upcoming results from Lipid17 have been mentioned in the blog:
    http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1522151331118\#c5876812436739342658} \\
  \todo{CHARMM36 results for this plot would be highly useful.}
\end{figure*}
In line with the trend observed for bilayers in the absence of negatively
charged lipids \cite{??}, the POPC headgroup order parameter response to lithium ions is
significantly larger than to potassium, indicating stronger binding of lithium
ions to lipid bilayers. The unique nature of lithium is also evident in the
response of PS headgroup order parameters to bounds ions, which decrease
with the addition of lithium or divalent cations but increase with the
addition of sodium or potassium. The headgroup order parameter response
to the added NaCl in the Berger model is similar to the experiments of LiCl,
except for the $\alpha$ carbon of POPS for which the change is overestimated
with the order of magnitude. The results suggests that the binding affinity
of sodium is overestimated by the Berger model, in line with the previous results \cite{catte16},
and that the response of $\alpha$ carbon of POPS to the added sodium
is significantly overestimated. The response of POPC to the added potassium
in MacRog model is only slightly overestimated, in contrast to the response
of POPS where order of magnitude overestimations of reposonse are observed. 
In conclusion, it seems that the response of PS headgroup to bound cations
is not correctly captured in the tested models.
\todo{Finish the discussion when more models are available}

>>>>

\subsection{Counterion binding to lipid bilayers containing PS lipids}
Membranes containing PS lipids are always accomppanied with counterions, which
modulate electrostatic interactions between lipids and other biomolecules. 
Counterions are also suggested screen the repulsion between charged lipid headgroups 
in MD simulations and reduce the area per lipid of PS bilayers to be smaller than in PC
bilayers~\cite{pandit02,mukhopadhyay04,pedersen06}. 
The counterion density profiles along membrane normal 
show significant differences between simulation models (Fig. \ref{NAdensPOPS}).
The strongest counterion binding, i.e., the lowest concentrations in bulk water,
are observed in MacRog, Berger and Lipid17/JC simulations.
CHARMM36, CHARMM36ua and Gromos-CKP models exhibit two local maxima in counterion
density, while a single maxima is observed in the other models. 
\todo{More detailed discussion may be possible after comparing monovalent ion binding
to bilayers between CHARMM simulations and experiments.}
Area per lipid is in agreement with experiments \cite{pan14} only
in the Gromos-CKP models, while other models give significantly lower values (Fig. \ref{NAdensPOPS}).
The difference cannot be explained by the electrostatic screening of the headgroup repulsion due to 
counterion binding because CHARMM36, CHARMM36ua and Slipid models give
smaller area per lipid than Gromos-CKP models with similar counterion binding affinity.
%The proposed trend is observed for Lipid17 model,
%for which the Joung-Cheatham ions give higher affinity and smaller area
%per molecule. However, the trend is not observed when comparing accross different
%force fields. For example, CHARMM36ua and MacRog give similar area per molecule
%but binding affinity is significantly higher in MacRog. 
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/NAdensPOPS.eps}
  \caption{\label{NAdensPOPS}
    Counterion densities of POPS lipid bilayer along the membrane normal from
    simulations with different force fields.
  }
\end{figure}

To evaluate counterion binding in different simulation against experimental data~\cite{roux90},
we plot the headgroup order parameters measured from POPC:POPS 5:1 mixture
as a function of function of different monovalent ions added to the buffer (Fig. \ref{PSresponseTONaCl}). 
\begin{figure*}[]
  \centering
  \includegraphics[width=18.0cm]{../Figs/CHANGESwithMONVALENTwithPS.eps}
  \caption{\label{PSresponseTONaCl}
    Changes of the PC (left) and PS (right) headgroup order parameters as a function of
    added NaCl, KCl and LiCl from POPC:POPS (5:1) mixture. The experimental data is from Ref. \citenum{roux90}.
    The values from counterion-only systems are set as a zero point of y-axis.
    To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
    in PS headgroup (bottom, right), the y-axis is tranferred with the same value for both order parameters such that the lower order
    parameter value is at zero.
  }
  \todo{The upcoming results from Lipid17 have been mentioned in the blog:
    http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1522151331118\#c5876812436739342658} \\
  \todo{CHARMM36 results for this plot would be highly useful.}
\end{figure*}
In line with the trend observed for bilayers in the absence of negatively
charged lipids \cite{??}, the POPC headgroup order parameter response to lithium ions is
significantly larger than to potassium, indicating stronger binding of lithium
ions to lipid bilayers. The unique nature of lithium is also evident in the
response of PS headgroup order parameters to bounds ions, which decrease
with the addition of lithium or divalent cations but increase with the
addition of sodium or potassium. The headgroup order parameter response
to the added NaCl in the Berger model is similar to the experiments of LiCl,
except for the $\alpha$ carbon of POPS for which the change is overestimated
with the order of magnitude. The results suggests that the binding affinity
of sodium is overestimated by the Berger model, in line with the previous results \cite{catte16},
and that the response of $\alpha$ carbon of POPS to the added sodium
is significantly overestimated. The response of POPC to the added potassium
in MacRog model is only slightly overestimated, in contrast to the response
of POPS where order of magnitude overestimations of reposonse are observed. 
In conclusion, it seems that the response of PS headgroup to bound cations
is not correctly captured in the tested models.
\todo{Finish the discussion when more models are available}
